---
title: "NHANES Workshop Demo Code"
author: "TTS Research Technology"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Packages

Install packages once if needed:

```{r install-packages, eval=FALSE}
install.packages("tidyverse")
install.packages("haven")
```

Load packages for today:

```{r packages}
library(tidyverse)
library(haven)
```

# Paths

This R Markdown file lives in the `code` folder. The `..` below means "go up one folder" to the main workshop folder.

```{r define-paths}
relative_path <- "../data/raw"
processed_path <- "../data/processed"
figure_path <- "../outputs/figures"

relative_path
```

Use `file.path()` instead of typing a long path by hand.

```{r file-path-example}
file.path(relative_path, "DEMO_J.xpt")
```

# Import NHANES Files

The workshop uses three NHANES 2017-2018 files:

- `DEMO_J.xpt`: demographics
- `BMX_J.xpt`: body measures, including BMI
- `DR1TOT_J.xpt`: Day 1 dietary intake

```{r import-xpt}
demo_data <- read_xpt(file.path(relative_path, "DEMO_J.xpt"))
bmx_data <- read_xpt(file.path(relative_path, "BMX_J.xpt"))
diet_data <- read_xpt(file.path(relative_path, "DR1TOT_J.xpt"))
```

Check that each object imported correctly.

```{r inspect-data}
head(demo_data, n = 5)
head(bmx_data, n = 5)
head(diet_data, n = 5)
```

Use `glimpse()` when you want a compact list of variable names and types. To keep the handout readable, this example shows only a few demographic variables.

```{r glimpse-data}
demo_data %>%
  select(SEQN, RIDAGEYR, RIAGENDR, WTINT2YR, WTMEC2YR, SDMVSTRA, SDMVPSU) %>%
  glimpse()
```

# Check the Join Key

Most participant-level NHANES files use `SEQN` as the respondent identifier. Before joining, check whether each file has one row per `SEQN`.

```{r check-duplicates}
demo_data %>%
  count(SEQN) %>%
  filter(n > 1)

bmx_data %>%
  count(SEQN) %>%
  filter(n > 1)

diet_data %>%
  count(SEQN) %>%
  filter(n > 1)
```

If no rows are printed, this demo has no duplicate `SEQN` values in these files.

# Select Variables

Keep the variables used later for joining, plotting, and missing-value checks.

```{r select-vars}
demo_small <- demo_data %>%
  select(SEQN, RIDAGEYR, RIAGENDR)

bmx_small <- bmx_data %>%
  select(SEQN, BMXBMI)

diet_small <- diet_data %>%
  select(SEQN, DR1TKCAL)
```

Check missing values before the join:

```{r missing-before-join}
missing_before_join <- bind_rows(
  demo_small %>%
    summarize(across(everything(), ~ sum(is.na(.)))) %>%
    mutate(file = "DEMO_J"),
  bmx_small %>%
    summarize(across(everything(), ~ sum(is.na(.)))) %>%
    mutate(file = "BMX_J"),
  diet_small %>%
    summarize(across(everything(), ~ sum(is.na(.)))) %>%
    mutate(file = "DR1TOT_J")
) %>%
  relocate(file)

missing_before_join
```

# Join Tables

Use `left_join()` to add columns from one participant-level file to another.

```{r join-tables}
nhanes_joined <- demo_small %>%
  left_join(bmx_small, by = "SEQN") %>%
  left_join(diet_small, by = "SEQN")

glimpse(nhanes_joined)
```

Because we started with `demo_small` and used `left_join()`, the row count should match the demographic file.

```{r check-joined-dimensions}
row_counts <- tibble(
  table = c("demo_small", "bmx_small", "diet_small", "nhanes_joined"),
  rows = c(
    nrow(demo_small),
    nrow(bmx_small),
    nrow(diet_small),
    nrow(nhanes_joined)
  )
)

row_counts
```

After a left join, variables from unmatched files become `NA`.

```{r missing-after-join}
nhanes_joined %>%
  summarize(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_n"
  )
```

# Save the Joined Dataset

Save a CSV copy in `data/processed`.

```{r save-processed}
write_csv(nhanes_joined, file.path(processed_path, "nhanes_joined_demo.csv"))
```

# Recode Variables

NHANES often stores categories as numbers. For this example, `RIAGENDR == 1` means Male and `RIAGENDR == 2` means Female.

The BMI groups below are analyst-created categories for teaching and visualization. The fixed cut points shown here are adult BMI cut points.

```{r recode}
nhanes_clean <- nhanes_joined %>%
  mutate(
    sex = case_when(
      RIAGENDR == 1 ~ "Male",
      RIAGENDR == 2 ~ "Female",
      TRUE ~ NA_character_
    ),
    bmi_group = case_when(
      is.na(BMXBMI) ~ NA_character_,
      BMXBMI < 18.5 ~ "Underweight",
      BMXBMI >= 18.5 & BMXBMI < 25 ~ "Healthy weight",
      BMXBMI >= 25 & BMXBMI < 30 ~ "Overweight",
      BMXBMI >= 30 ~ "Obesity"
    ),
    bmi_group = factor(
      bmi_group,
      levels = c("Underweight", "Healthy weight", "Overweight", "Obesity")
    ),
    age_group = case_when(
      is.na(RIDAGEYR) ~ NA_character_,
      RIDAGEYR < 40 ~ "Younger than 40",
      RIDAGEYR >= 40 & RIDAGEYR < 60 ~ "40 to 59",
      RIDAGEYR >= 60 ~ "60 or older"
    )
  )

nhanes_clean %>%
  select(SEQN, RIDAGEYR, age_group, sex, BMXBMI, bmi_group, DR1TKCAL)
```

Use `glimpse()` again to see how the new character and factor variables differ from the original numeric variables.

```{r glimpse-after-recode}
nhanes_clean %>%
  select(SEQN, RIDAGEYR, age_group, sex, BMXBMI, bmi_group, DR1TKCAL) %>%
  glimpse()
```

# Plot 1: BMI by Age

In `ggplot2`, start with a dataset, map variables to visual features, then add layers.

```{r plot-bmi-age}
bmi_age_plot <- ggplot(
  data = nhanes_clean,
  mapping = aes(x = RIDAGEYR, y = BMXBMI, color = sex)
) +
  geom_point(size = 2.5, alpha = 0.75) +
  labs(
    title = "BMI by age in the NHANES workshop sample",
    x = "Age in years",
    y = "Body mass index",
    color = "Sex"
  ) +
  theme_minimal()

bmi_age_plot
```

Save the figure:

```{r save-bmi-age}
ggsave(
  filename = file.path(figure_path, "bmi_by_age.png"),
  plot = bmi_age_plot,
  width = 7,
  height = 4.5,
  dpi = 300
)
```

# Plot 2: Energy Intake by BMI Group

This plot uses `BMXBMI` from the body-measures file and `DR1TKCAL` from the dietary file.

```{r plot-energy-bmi}
energy_plot <- ggplot(
  data = nhanes_clean,
  mapping = aes(x = bmi_group, y = DR1TKCAL)
) +
  geom_violin(fill = "#D8ECF8", color = "#2C5D75", trim = FALSE, na.rm = TRUE) +
  geom_boxplot(width = 0.14, outlier.shape = NA, fill = "white", na.rm = TRUE) +
  labs(
    title = "Day 1 energy intake by BMI group",
    x = "BMI group",
    y = "Day 1 total energy intake, kcal"
  ) +
  theme_minimal()

energy_plot
```

Save the figure:

```{r save-energy-bmi}
ggsave(
  filename = file.path(figure_path, "energy_by_bmi_group.png"),
  plot = energy_plot,
  width = 7,
  height = 4.5,
  dpi = 300
)
```

Important interpretation reminder: this is an unweighted plot. It is useful for learning data structure and visualization, but it is not a population estimate.

# Practice Plot

Modify the first plot so it shows BMI group and sex.

```{r practice-plot}
practice_plot <- ggplot(
  data = nhanes_clean,
  mapping = aes(x = RIDAGEYR, y = BMXBMI, color = bmi_group)
) +
  geom_point(size = 2.5, alpha = 0.75) +
  facet_wrap(~ sex) +
  labs(
    title = "BMI by age, BMI group, and sex",
    x = "Age in years",
    y = "Body mass index",
    color = "BMI group"
  ) +
  theme_minimal()

practice_plot

ggsave(
  filename = file.path(figure_path, "practice_bmi_by_age.png"),
  plot = practice_plot,
  width = 8,
  height = 4.5,
  dpi = 300
)
```

Write two notes after running your plot:

1. One pattern you notice.
2. One reason this plot should not be treated as a final NHANES estimate.
